Supplement to ``{\mytitle}''

### 2022-12-14 si/si.Rnw
Minor edits

<<debug,echo=F>>=
run_level <- 1
@

@@

<<Load lee Model1 Params, echo=FALSE>>=
h1_epi <- haiti1()
RL_dir <- paste0("run_level_", run_level, "/")
@

<<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
N_LEE_SIMS <- 20

@@

<<echo=FALSE, include=FALSE, message=FALSE>>=
h1_epi_evals <- readRDS(paste0("../model1/", RL_dir, "lee1_epi_evals.rds"))
h1_end_evals <- readRDS(paste0("../model1/", RL_dir, "lee1_end_evals.rds"))

@@

<<Model3Replication, echo=FALSE, cache=TRUE>>=
NSIM <- switch(run_level, 50, 200, 500)

@@

lee3 <- lee3_spatPomp()
lee3_sims <- simulate(lee3, nsim = NSIM, seed = 321, format = 'data.frame')

@@

>>>>

@@

<<Simulate Lee Model 1, echo=FALSE, cache=TRUE>>=
h1_epi <- haiti1()
N_LEE_SIMS <- 20

@@

<<echo=FALSE, include=FALSE, message=FALSE>>=
h1_epi_evals <- readRDS("../model1/run_level_3/lee1_epi_evals.rds")
h1_end_evals <- readRDS("../model1/run_level_3/lee1_end_evals.rds")

@@

<<Model3Replication, echo=FALSE, cache=TRUE>>=
NSIM <- 500

@@

lee3 <- lee3_spatPomp()
lee3_sims <- bake("../model3/si_lee3Sims.rds", {
  simulate(lee3, nsim = NSIM, seed = 321, format = 'data.frame')
})

@@

\section{Calibrating Model~3 to observed cases}

<<CalibrateOrig3, echo=FALSE>>=
set.seed(8636712)

# Set Run Level for debug, timing, and full computation
RUN_LEVEL = 2

# Create vectors for the unit and shared parameters
unit_specific_names <- c("betaB", "foi_add")
shared_param_names <- c(
  "mu_B", "XthetaA", "thetaI", "lambdaR", "r", "std_W",
  "epsilon", "k"
)
est_param_names <- c(
  unit_specific_names, shared_param_names
)

# Add unit numbers to each parameter
est_param_names_expanded <- paste0(rep(est_param_names, each = 10), 1:10)

set_rw_1 <- function(x) {
  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
    return(0.01)
  } else if (x %in% paste0("k", 1:10)) {
    return(0.01)
  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
    return(0.01)
  } else {
    return(0)
  }
}

set_rw_2 <- function(x) {
  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
    return(0.003)
  } else if (x %in% paste0("k", 1:10)) {
    return(0.003)
  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
    return(0.003)
  } else {
    return(0)
  }
}

set_rw_3 <- function(x) {
  if (gsub("[[:digit:]]+$", "", x) %in% shared_param_names) {
    return(0.0025)
  } else if (x %in% paste0("k", 1:10)) {
    return(0.0025)
  } else if (grepl('^foi_add[[:digit:]]+$', x) || grepl('^betaB[[:digit:]]+$', x)) {
    return(0.0025)
  } else {
    return(0)
  }
}

# First Search RW
reg_rw_1.sd <- lapply(est_param_names_expanded, set_rw_1)
names(reg_rw_1.sd) <- est_param_names_expanded
chol_rw_1.sd <- do.call(rw.sd, reg_rw_1.sd)

# Second Search RW
reg_rw_2.sd <- lapply(est_param_names_expanded, set_rw_2)
names(reg_rw_2.sd) <- est_param_names_expanded
chol_rw_2.sd <- do.call(rw.sd, reg_rw_2.sd)

# Third Search RW
reg_rw_3.sd <- lapply(est_param_names_expanded, set_rw_3)
names(reg_rw_3.sd) <- est_param_names_expanded
chol_rw_3.sd <- do.call(rw.sd, reg_rw_3.sd)

# Search hyper-parameters for each run-level and search
SEARCH1 = list(
  NBPF = 100,
  NP = 1000,
  SPAT_REGRESSION = 0.05,
  NREPS = 72,
  NP_EVAL = 1000,
  NREPS_EVAL = 8,
  RW_SD = chol_rw_1.sd,
  COOLING = 0.5,
  KEEP_TRACES = TRUE,
  KEEP_LIK_MAT = TRUE
)

SEARCH2 = list(
  TOP_N = 9,
  NBPF = 100,
  NP = 1000,
  SPAT_REGRESSION = 0.05,
  NREPS = 4,
  NP_EVAL = 1000,
  NREPS_EVAL = 9,
  RW_SD = chol_rw_2.sd,
  COOLING = 0.5,
  KEEP_TRACES = TRUE,
  KEEP_LIK_MAT = TRUE
)

SEARCH3 = list(
  TOP_N = 6,
  NBPF = 100,
  NP = 1000,
  SPAT_REGRESSION = 0.05,
  NREPS = 6,
  NP_EVAL = 1000,
  NREPS_EVAL = 9,
  RW_SD = chol_rw_3.sd,
  COOLING = 0.5,
  KEEP_TRACES = TRUE,
  KEEP_LIK_MAT = TRUE
)

N_SEARCHES <- 3L

save_file <- "../model3/run_level_2/fit_orig_model.rds"

# TODO: replace with bake
haiti3_orig_fit <- readRDS(save_file)

# haiti3_orig_fit <- bake(
#   file = save_file, {
#     fit_coupled_haiti3(
#       nsearches = N_SEARCHES,
#       search1 = SEARCH1,
#       search2 = SEARCH2,
#       search3 = SEARCH3,
#       search_rho = FALSE,
#       search_gamma = FALSE,
#       ncores = cores
#     )
#   }
# )

best_m <- which.max(haiti3_orig_fit$search3$logLiks$logLik)

best_params <- haiti3_orig_fit$search3$params[best_m, ]

all_liks <- as.data.frame(haiti3_orig_fit$search3$likMat)
all_liks$which <- rep(1:36, each = 9)
best_mod <- filter(all_liks, which == best_m)

unitLikesM3 <- best_mod %>%
  pivot_longer(
    cols = -which,
    names_to = "unit",
    names_prefix = "V",
    values_to = "logLik"
  ) %>%
  select(-which) %>%
  group_by(unit) %>%
  summarize(
    ll_est = logmeanexp(logLik),
    ll_se = logmeanexp(logLik, se = TRUE)[2]
  ) %>%
  ungroup() %>%
  mutate(unit = as.numeric(unit)) %>%
  arrange(unit)
@

<<M3_bench, echo=FALSE, message=FALSE, include=FALSE>>=
MODEL3_CASES <- haitiCholera %>%
  dplyr::rename(
    date = date_saturday, Grande_Anse = Grand.Anse,
    Nord_Est = Nord.Est, Nord_Ouest = Nord.Ouest,
    Sud_Est = Sud.Est
  ) %>%
  dplyr::mutate(date = as.Date(date)) %>%
  dplyr::select(-report) %>%
  dplyr::mutate(time_diff = date - lag(date))

jumps <- which(MODEL3_CASES$time_diff > lubridate::make_difftime(day = 7))

temp <- MODEL3_CASES
for (j in 1:length(jumps)) {
  temp <- rbind(
    temp[1:(jumps[j] + j - 2), ],
    rep(NA, ncol(temp)),
    temp[(jumps[j] + j - 1):nrow(temp), ]
  )
}

mod3_cases <- temp %>%
  dplyr::select(-time_diff)

mod3_departements <-
  c(
    'Artibonite',
    'Centre',
    'Grande_Anse',
    'Nippes',
    'Nord',
    'Nord-Est',
    'Nord-Ouest',
    'Ouest',
    'Sud',
    'Sud-Est'
  )

m3_benchmark <- c()
for (dep in mod3_departements) {

  # Get department cases
  cases <- as.numeric(mod3_cases[, gsub("-", "_", dep)])

  # Fit model on natural scale of cases
  m3_a201 <- arima(cases, order = c(2, 0, 1))

  # Fit model on log-scale of cases
  m3_log_a201 <- arima(log(cases + 1), order = c(2, 0, 1))

  m3_benchmark <- c(m3_benchmark, max(
    m3_log_a201$loglik - sum(log(cases + 1), na.rm = TRUE),
    m3_a201$loglik
  )
  )
}

rm(cases, m3_a201, m3_log_a201, mod3_departements, dep,
   mod3_cases, temp, jumps, MODEL3_CASES, j)

unitLikesARMA <- data.frame(
  unit = 1:10,
  ll = m3_benchmark
) %>%
  as_tibble()
@

<<combineARMA_M3, echo=FALSE>>=
h3 <- haiti3_spatPomp()
deps <- h3@unit_names
ARMA_M3_joined <- dplyr::left_join(
  x = unitLikesM3,
  y = unitLikesARMA,
  by = "unit"
) %>%
  select(-ll_se) %>%
  mutate(unit = as.character(unit)) %>%
  pivot_longer(
    cols = -unit,
    names_to = "type",
    values_to = "ll"
  ) %>%
  mutate(unit = paste0(unit, ": ", deps[as.numeric(unit)]))
@

<<M3bpf, echo=FALSE, message=FALSE, include=FALSE>>=

coef(h3)[names(best_params)] <- best_params

registerDoRNG(40765101)

bpf_out <- bake(
  file = "../model3/run_level_3/si_bpf.rds", {
    h3 %>% bpfilter(
      Np = 2000,
      block_size = 1,
      save_states = TRUE
    )
  }
)

In this section, we provide more detail on the process that was used to estimate the coefficients used in Model~3.
In particular, we discuss why we decided to include additional model parameters that were not considered by \cite{lee20} (parameters related to initial conditions and Hurricane Matthew).
To calibrate this model, we used the iterated block particle filter (IBPF) method of \cite{ionides22}.
Due to the novelty of this algorithm, there does not exist published examples the IBPF algorithm used for data analysis outside of the papers in which the algorithm was introduced \citep{ning21ibpf,ionides22}, which at least partially motivates the creation of this section.

\cite{lee20} were only able to estimate model parameters to a simplified version of Model~3 on a subset of the available data, as no method existed at the time of their publication to fit a fully coupled meta-population model to disease incidence data.
Building on their results, we fit the fully coupled version of Model~3 to all available data.
Our preliminary results (not shown) suggest that a single search for the MLE using the IBPF algorithm is insufficient;
this result is consistent with the findings of \cite{ionides22}, who used a sequence of successive searches for the MLE in order to properly maximize the model likelihood.
We therefore use a similar approach by performing three successive searches for the MLE.
The first search is performed by obtaining initial values for the parameters by uniformly sampling values from a predefined hypercube.
Subsequent searches use the resulting parameter values that correspond the largest model likelihoods.

We use the technique described above to fit the fully coupled version of Model~3 proposed by \cite{lee20} to all observed cholera data.
The maximum likelihood we obtained with the model was $\Sexpr{round(max(haiti3_orig_fit$search3$logLiks$logLik))}$, which is higher than the log-ARMA benchmark ($\Sexpr{round(sum(unitLikesARMA$ll))}$).
For meta-population models, it is worth considering how well the model fits the data to each spatial unit.
This can be done by looking at conditional log-likelihoods, which is part of the output of the \texttt{bpfilter} algorithm in the \texttt{spatPomp} package.
The likelihoods for each department, compared to the corresponding log-ARMA benchmark, are displayed in Fig.~\ref{fig:h3UnitLikes}.
The figure demonstrates that while the log-likelihood of the fitted model outperforms the log-ARMA benchmark, the model has lower likelihoods than the benchmark for some departments.
This result warrants further investigation of the model fit.

\begin{figure}[ht]
<<h3UnitLikes, fig.height=3, fig.width=5.5, echo=FALSE>>=
ggplot(ARMA_M3_joined, aes(x = ll, y = reorder(unit, ll))) +
  geom_point(aes(col = type)) +
  ylab("Unit") +
  xlab("Log Likelihood") +
  scale_color_manual(
    labels = c("ll" = "ARMA", "ll_est" = "Model 3"),
    values = c("ll" = "#1b9e77", "ll_est" = "#d95f02")
  ) +
  theme(legend.title = element_blank(),
        axis.title.y = element_blank())
@
\caption{\label{fig:h3UnitLikes}Log-likelihoods of Model~3 for each department compared to the corresponding log-ARMA benchmark.}
\end{figure}

To investigate the goodness-of-fit of Model~3 further and attempt to determine why the model is underperforming in some departments, we plot the conditional log likelihoods of each observation given the previous observations for each department in Fig.~\ref{fig:condLL}.
This figure reveals that the fitted model has difficult time explaining certain features of the data.
For example, many departments (in particular Sud and Ouest) have observations with lower conditional log-likelihoods near October 2016 than at other time points.
Further investigation of the model output reveals that the model is struggling to explain the sudden surge in cholera cases that occurred at this time, which coincides with the time that Hurricane Matthew struck Haiti.
While the model does include a mechanism to account for increased cholera transmission due to large rainfall events, the mechanism does not appear to be sufficient to capture the damaging effects of the hurricane, which was most impactful in the the Sud and Ouest departments \jwc{TODO: CITE}.
This result led us to include parameters that would increase the transmission between environmental cholera and humans for Sud and Ouest \jwc{TODO: just check these are the correct departments}.
Other \jwc{TODO: IVPs}.

\begin{figure}[ht]
<<CondLLFig, echo=FALSE, message=FALSE, fig.height=3>>=
deps[grepl("_", deps)] <- gsub("_", "-", deps[grepl("_", deps)])
deps[deps == "Grande-Anse"] <- "Grand\'Anse"
names(deps) <- as.character(1:10)

myLabeller <- as_labeller(deps)

t(bpf_out@block.cond.loglik) %>%
  as.data.frame() %>%
  mutate(t = bpf_out@times) %>%
  pivot_longer(
    cols = -t,
    values_to = 'likelihood',
    names_prefix = "V",
    names_to = 'Unit'
  ) %>%
  mutate(date = as.Date(lubridate::date_decimal(t))) %>%
  ggplot(aes(x = date, y = likelihood)) +
  geom_line() +
  geom_point() +
  facet_wrap(~Unit, labeller = myLabeller, scales = 'free_y',
             nrow = 2) +
  theme_bw() +
  theme(axis.title.x = element_blank()) +
  ylab("Conditional log-likelihoods")
@
\caption{\label{fig:condLL}Conditional log-likelihoods of Model~3 prior to the inclusion of parameters for initial values and Hurricane Matthew.}
\end{figure}
